---
title: "Weekly Lab Homework Assignment: Interaction Effects"
author: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

*Replace "Your Name" with your actual name.*

## Instructions

In this lab assignment, you will practice interpreting interactions by visualizing them with different types of graphs. You'll work with simulated datasets and explore interactions between categorical x categorical, linear x linear, and categorical x linear variables.

Please follow the instructions for each exercise, and use `ggplot2` for all visualizations.

### Exercise 1: Categorical x Categorical Interaction

**Task**: Use the dataset with two categorical variables and one outcome variable. Fit a model with a categorical x categorical interaction, and visualize the interaction using a bar graph with error bars.

1.  Use the dataset with two categorical variables: `Education_Level` (e.g., "High School", "College") and `Job_Type` (e.g., "Office", "Field"), and an outcome variable `Job_Satisfaction`.
2.  Fit a linear model including the interaction between `Education_Level` and `Job_Type` on `Job_Satisfaction`.
3.  Create a bar graph with error bars to visualize the interaction.
4.  Interpret the interaction term and the graph.

```{r, warning = F, message = F}
library(ggplot2)
library(dplyr)

# Simulate data
set.seed(123)
Education_Level <- factor(rep(c("High School", "College"), each = 50))
Job_Type <- factor(rep(c("Office", "Field"), each = 25, times = 2))
Job_Satisfaction <- ifelse(Education_Level == "College", 
                           70 + 10 * (Job_Type == "Office"), 
                           60 + 5 * (Job_Type == "Field")) + rnorm(100, sd = 5)

data1 <- data.frame(Education_Level, Job_Type, Job_Satisfaction)

# Calculate means and standard errors with correct handling
means1 <- data1 %>%
  group_by(Education_Level, Job_Type) %>%
  summarise(
    Job_Satisfaction_Mean = mean(Job_Satisfaction),
    SE = (sd(Job_Satisfaction) / sqrt(n())),  # Calculate SE 
    lower = Job_Satisfaction_Mean - 1.96 * SE,  # Lower bound of the confidence interval
    upper = Job_Satisfaction_Mean + 1.96 * SE   # Upper bound of the confidence interval
  )

# Check the calculated values
print(means1)

# Plot the bar graph with error bars
ggplot(means1, aes(x = Job_Type, y = Job_Satisfaction_Mean, fill = Education_Level)) +
  geom_bar(stat = "identity", position = position_dodge(0.9), width = 0.7) +
  geom_errorbar(aes(ymin = lower, ymax = upper),
                width = 0.2, position = position_dodge(0.9)) +
  labs(title = "Interaction between Education Level and Job Type on Job Satisfaction",
       x = "Job Type", y = "Mean Job Satisfaction with 95% CI") +
  scale_fill_manual(values = c("High School" = "lightblue", "College" = "lightgreen")) +
  theme_minimal()

```

**Interpretation**: The interaction term will tell you how the effect of job type on job satisfaction changes depending on the education level. The bar graph will visualize these differences.

### Exercise 2: Linear x Linear Interaction

**Task**: Use the dataset with two continuous variables and one outcome variable. Fit a model with a linear x linear interaction, and visualize the interaction using a 2D plot with a median split of one variable.

1.  Use the two continious predictors from the dataset: `Age` and `Weekly_Hours_Worked`, and an outcome variable `Income`.
2.  Fit a linear model including the interaction between `Age` and `Weekly_Hours_Worked`.
3.  Create a 2D plot using a median split of `Weekly_Hours_Worked` to visualize the interaction.
4.  Interpret the interaction term and the graph.

```{r}
# Simulate data
set.seed(123)
Age <- rnorm(100, mean = 40, sd = 10)
Weekly_Hours_Worked <- rnorm(100, mean = 40, sd = 5)
Income <- 30000 + 500 * Age + 1000 * Weekly_Hours_Worked + 50 * Age * Weekly_Hours_Worked + rnorm(100, sd = 5000)

data2 <- data.frame(Age, Weekly_Hours_Worked, Income)

# Fit the model
model2 <- lm(Income ~ Age * Weekly_Hours_Worked, data = data2)

# Summary of the model
summary(model2)

# Create a median split for Weekly_Hours_Worked
data2 <- data2 %>%
  mutate(Hours_Level = ifelse(Weekly_Hours_Worked > median(Weekly_Hours_Worked), "High", "Low"))

# Plot the interaction
ggplot(data2, aes(x = Age, y = Income, color = Hours_Level)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Effect of Age on Income by Weekly Hours Worked (Median Split)",
       x = "Age", y = "Income") +
  scale_color_manual(values = c("High" = "blue", "Low" = "red")) +
  theme_minimal()
```

**Interpretation**: The interaction term will tell you how the effect of age on income changes depending on the level of weekly hours worked. The plot will visualize how these relationships differ.

### Exercise 3: Categorical x Linear Interaction

**Task**: Use the simulated dataset with one categorical variable and one continuous variable as predictors. Fit a model with a categorical x linear interaction, and visualize the interaction using an interaction plot.

1.  Use the dataset with the two predictos: one categorical variable `Gender` and one continuous variable `Study_Hours`, and an outcome variable `Test_Score`.
2.  Fit a linear model including the interaction between `Gender` and `Study_Hours`.
3.  Create an interaction plot using `ggplot2` to visualize the interaction.
4.  Interpret the interaction term and the graph.

```{r}
# Simulate data
set.seed(123)
Gender <- factor(rep(c("Male", "Female"), each = 50))
Study_Hours <- rnorm(100, mean = 5, sd = 2)
Test_Score <- 60 + 10 * (Gender == "Female") + 5 * Study_Hours + 5 * (Gender == "Female") * Study_Hours + rnorm(100, sd = 5)

data3 <- data.frame(Gender, Study_Hours, Test_Score)

# Fit the model
model3 <- lm(Test_Score ~ Gender * Study_Hours, data = data3)

# Summary of the model
summary(model3)

# Plot the interaction
ggplot(data3, aes(x = Study_Hours, y = Test_Score, color = Gender)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Interaction between Gender and Study Hours on Test Scores",
       x = "Study Hours", y = "Test Score") +
  scale_color_manual(values = c("Male" = "lightblue", "Female" = "lightcoral")) +
  theme_minimal()
```

**Interpretation**: The interaction term will tell you how the effect of study hours on test scores differs between males and females. The plot will visualize how these relationships differ.

### Exercise 4: Visualizing Multivariate Interactions

**Task**: Given a multivariate dataset, create different types of graphs to visualize interactions and discuss which type of graph is most appropriate.

1.  Use the simulated multivariate dataset with both continuous and categorical variables.
2.  Create a graph for each type of interaction (categorical x categorical, categorical x continuous, continuous x continuous) . Use a median split for the continuous x continuous interaction.
3.  Discuss which type of graph best represents the interactions in your data and why.

```{r, warning = F, message = F}
# Simulate data
set.seed(123)
Age <- rnorm(100, mean = 35, sd = 8)
Experience <- rnorm(100, mean = 10, sd = 5)
Gender <- factor(rep(c("Male", "Female"), each = 50))
Job_Type <- factor(rep(c("Office", "Field"), each = 25, times = 2))
Salary <- 30000 + 1000 * Age + 2000 * Experience + 150 * Age * Experience + 
  5000 * (Gender == "Female") + rnorm(100, sd = 5000)

data4 <- data.frame(Age, Experience, Gender, Job_Type, Salary)

# Categorical x Categorical Interaction (Bar graph with means and SE)
# Summarizing means and standard errors
means4 <- data4 %>%
  group_by(Gender, Job_Type) %>%
  summarise(
    Mean_Salary = mean(Salary),
    SE_Salary = sd(Salary) / sqrt(n())
  )

# Plot the bar graph with error bars
ggplot(means4, aes(x = Job_Type, y = Mean_Salary, fill = Gender)) +
  geom_bar(stat = "identity", position = position_dodge(0.9), width = 0.7) +
  geom_errorbar(aes(ymin = Mean_Salary - 1.96 * SE_Salary, ymax = Mean_Salary + 1.96 * SE_Salary),
                width = 0.2, position = position_dodge(0.9)) +
  labs(title = "Interaction between Gender and Job Type on Salary",
       x = "Job Type", y = "Mean Salary with 95% CI") +
  scale_fill_manual(values = c("Male" = "lightblue", "Female" = "lightcoral")) +
  theme_minimal()

# Interaction plot (categorical x continuous)
ggplot(data4, aes(x = Age, y = Salary, color = Gender)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Interaction between Gender and Age on Salary",
       x = "Age", y = "Salary") +
  scale_color_manual(values = c("Male" = "blue", "Female" = "red")) +
  theme_minimal()

# Faceted plot (categorical x continuous)
ggplot(data4, aes(x = Experience, y = Salary)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Effect of Experience on Salary by Gender",
       x = "Experience", y = "Salary") +
  facet_wrap(~ Gender) +
  theme_minimal()

# Median split for Experience to visualize the linear x linear interaction
data4 <- data4 %>%
  mutate(Experience_Level = ifelse(Experience > median(Experience), "High Experience", "Low Experience"))

# Plot the interaction using the median split
ggplot(data4, aes(x = Age, y = Salary, color = Experience_Level)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Effect of Age on Salary by Experience Level (Median Split)",
       x = "Age", y = "Salary") +
  scale_color_manual(values = c("High Experience" = "green", "Low Experience" = "orange")) +
  theme_minimal()
```

**Discussion**: Based on the graphs, decide which type best visualizes the interactions in your dataset and explain why. For example, faceted plots might be most effective for categorical x continuous interactions, while 3D surface plots are better for continuous x continuous interactions.

**Submission Instructions:**

Ensure to knit your document to PDF format, checking that all content is correctly displayed before submission. Submit this PDF to Canvas Assignments.
