---
title: "Weekly Lab Homework Assignment: Assessing Model Fit"
author: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

*Replace "Your Name" with your actual name.*

## Instructions

This lab assignment is designed to reinforce your understanding of assessing model fit using Chi-Square tests, R-squared, and F-tests. You will apply these concepts to different datasets, calculate relevant statistics, and interpret the results.

### Exercise 1: Chi-Square Goodness of Fit Test

**Objective**: Perform a Chi-Square goodness of fit test using the provided dataset. Interpret the Chi-Square statistic and p-value to determine whether the observed frequencies differ significantly from the expected frequencies.  

**Dataset**:  
- **Observed frequencies**: `observed <- c(20, 30, 50)` (e.g., three different responses to a new therapy)  
- **Expected frequencies**: `expected <- c(33.3, 33.3, 33.3)` (assuming equal preference for each response)  

**Tasks**:  
1. Use the `chisq.test()` function in R to perform the Chi-Square test.  
2. Include the code output in your .rmd file.  
3. Write a brief interpretation of the Chi-Square statistic and p-value.  

```{r}
# Observed and expected frequencies
observed <- c(20, 30, 50)
expected <- c(33.3, 33.3, 33.3)

# Perform Chi-Square test

```

### Exercise 2: Calculating R-Squared for a Linear Regression Model

**Objective**: Calculate the R-squared value for a linear regression model using the provided dataset. Interpret the R-squared value and discuss what it indicates about the model’s fit.  

**Dataset**:  
- **Independent variable**: `hours_practiced <- c(2, 3, 5, 7, 8, 10, 12, 14, 15)`  
- **Dependent variable**: `performance_score <- c(50, 55, 60, 65, 70, 75, 80, 85, 88)`  

**Tasks**:  
1. Fit a linear regression model using the `lm()` function in R.  
2. Calculate the R-squared value.  
3. Include the R output and your interpretation in the .rmd file.  

```{r}
# Data for regression analysis
hours_practiced <- c(2, 3, 5, 7, 8, 10, 12, 14, 15)
performance_score <- c(50, 55, 60, 65, 70, 75, 80, 85, 88)

# Fit linear regression model

# Calculate R-squared

```

### Exercise 3: Comparing Nested Models Using the F-Test

**Objective**: Compare two nested linear regression models using the F-test. Determine whether adding an additional predictor variable significantly improves the model. Interpret the F-statistic and p-value.  

**Dataset**:  
- **Predictor 1**: `study_hours <- c(2, 4, 6, 8, 10, 12, 14, 16, 18)`  
- **Predictor 2**: `part_time_job <- c(0, 1, 0, 1, 0, 1, 0, 1, 0)` (0 = no, 1 = yes)  
- **Dependent variable**: `exam_score <- c(55, 60, 62, 70, 72, 78, 80, 85, 87)`  

**Tasks**:  
1. Fit a simple linear regression model using `study_hours` as the predictor.  
2. Fit a more complex model using both `study_hours` and `part_time_job` as predictors.  
3. Perform an F-test using the `anova()` function to compare the models.  
4. Include the code, F-statistic, p-value, and your interpretation.  

```{r}
# Data for regression analysis
study_hours <- c(2, 4, 6, 8, 10, 12, 14, 16, 18)
part_time_job <- c(0, 1, 0, 1, 0, 1, 0, 1, 0)
exam_score <- c(55, 60, 62, 70, 72, 78, 80, 85, 87)

# Fit simple model
 

# Fit more complex model
 
# Perform F-test to compare models

```

### Exercise 4: Visualizing the F-Distribution

**Objective**: Visualize the F-distribution with different degrees of freedom in R. Discuss how the shape of the F-distribution affects the critical value used in the F-test.  

**Tasks**:  
1. Plot the F-distribution in R using different degrees of freedom.  
2. Observe how the shape of the F-distribution changes with different degrees of freedom.  
3. Include the plots and a brief discussion in your .rmd file.  

```{r}
# Load ggplot2
library(ggplot2)

# Define F-values
f_values <- seq(0, 4, length.out = 100)

# Plot F-distribution for different degrees of freedom

```

### Exercise 5: Comprehensive Analysis

**Objective**: Conduct a comprehensive analysis that includes a Chi-Square test, R-squared calculation, and an F-test. Use the provided dataset and write a summary report of your findings.  

**Dataset**:  
- **Observed frequencies**: `observed_frequencies <- c(25, 40, 35)`  
- **Expected frequencies**: `expected_frequencies <- c(33.3, 33.3, 33.3)`  
- **Independent variable**: `study_time <- c(5, 6, 7, 8, 9, 10, 11, 12, 13)`  
- **Dependent variable**: `stress_levels <- c(7, 8, 6, 7, 8, 9, 10, 11, 12)`  
- **Additional predictor**: `exercise_hours <- c(2, 3, 1, 4, 2, 3, 5, 6, 4)`  

**Tasks**:  
1. Perform a Chi-Square goodness of fit test using the observed and expected frequencies.  
2. Fit a regression model and calculate R-squared using `study_time` to predict `stress_levels`.  
3. Compare two models using the F-test by adding `exercise_hours` as an additional predictor.  
4. Include all code, output, and a summary interpretation in your .rmd file.  

*Chi-Square Test*:
```{r}
observed_frequencies <- c(25, 40, 35)
expected_frequencies <- c(33.3, 33.3, 33.3)


```

*R-Squared Calculation*:
```{r}
study_time <- c(5, 6, 7, 8, 9, 10, 11, 12, 13)
stress_levels <- c(7, 8, 6, 7, 8, 9, 10, 11, 12)


```

*F-Test*:
```{r}
exercise_hours <- c(2, 3, 1, 4, 2, 3, 5, 6, 4)

model1 <- lm(stress_levels ~ study_time)
model2 <- lm(stress_levels ~ study_time + exercise_hours)


```


**Submission Instructions:**

Ensure to knit your document to PDF format, checking that all content is correctly displayed before submission. Submit this PDF to Canvas Assignments.
