---
title: "Weekly Lab Homework Assignment: Measurement Errors in Psychological Research"
author: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

*Replace "Your Name" with your actual name.*

## Objective:

Reinforce your understanding of reliability, validity, and data collection errors through hands-on exercises using R. Apply statistical methods to simulate data, calculate various types of reliability and validity, and interpret the outcomes.

## Setup:

Prepare your RStudio environment by installing and loading the necessary packages:

```{r}
# Install and load required packages
if(!require(psych)) {
    install.packages("psych", dependencies=TRUE)
}
library(psych)
```

## Homework Exercises:

### Exercise 1: Test-Retest Reliability

Evaluate test-retest reliability for a psychological test administered twice.

**Task**:

1.  Simulate test scores for 50 participants at two time points.

2.  Calculate the Pearson correlation coefficient.

3.  Interpret the test-retest reliability of the scores.

```{r}
set.seed(2024)
scores_time1 <- rnorm(50, mean=100, sd=10)
scores_time2 <- scores_time1 + rnorm(50, mean=0, sd=5)

# Insert your code below
cor.test(scores_time1, scores_time2)
```

**Write your interpretation here**

### Exercise 2: Internal Consistency

Assess internal consistency of a psychological scale.

**Task**:

1.  Simulate responses for 100 participants on a 10-item scale.

2.  Calculate Cronbach's alpha.

```{r}
responses <- as.data.frame(matrix(rnorm(1000, mean=3, sd=1), ncol=10))

## Insert your code below
alpha(responses)

```

**Write your interpretation here**

#### Exercise 3: Inter-Rater Reliability

Measure agreement between two raters.

**Task**:

1.  Simulate two sets of ratings for 30 subjects.

2.  Calculate Cohen's Kappa.

```{r}
rater1 <- sample(1:5, 30, replace=TRUE)
rater2 <- sample(1:5, 30, replace=TRUE)

## Insert your code below
# Calculate Cohen’s Kappa for inter-rater reliability
cohen.kappa(matrix(c(rater1, rater2), ncol=2))

```

**Write your interpretation here**

#### Exercise 4: Convergent Validity

Investigate convergent validity of a new test.

**Task**:

1.  Simulate scores for a new test and an established test.

2.  Calculate the correlation to assess convergent validity.

```{r}
new_test_scores <- rnorm(100, mean=50, sd=10)
established_test_scores <- new_test_scores * 1.1 + rnorm(100, mean=0, sd=5)

## Insert your code below
# Calculate the correlation for convergent validity
cor(new_test_scores, established_test_scores)


```

**Write your interpretation here**

#### Exercise 5: Divergent Validity

Explore divergent validity by examining the relationship between measures of unrelated constructs.

**Task**:

1.  Simulate scores for a test of social anxiety and a test of mathematical ability.

2.  Calculate the correlation to evaluate divergent validity.

```{r}
social_anxiety_scores <- rnorm(100, mean=50, sd=10)
math_ability_scores <- rnorm(100, mean=50, sd=10)

# Assess divergent validity by calculating the correlation
## Insert your code below
# Assess divergent validity by calculating the correlation
cor(social_anxiety_scores, math_ability_scores)

```

**Write your interpretation here**

#### Submission Instructions:

Ensure to knit your document to PDF format, checking that all content is correctly displayed before submission. Submit this PDF to Canvas Assignments.
