---
title: "Weekly Lab Homework Assignment: Descriptive Statistics and Probability"
author: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

*Replace "Your Name" with your actual name.*

## Objective:

This lab assignment aims to reinforce your understanding of descriptive statistics, calculating probabilities, and identifying sample spaces. You will apply these concepts to practical problems using R.

## Homework Exercises:

### Exercise 1: Analyzing Descriptive Statistics

-   **Task**: Given a dataset of reaction times, calculate the mean, median, mode, variance, standard deviation, and identify any outliers.

-   **Dataset**: Use the following reaction times (in milliseconds) for the analysis: `c(250, 340, 295, 305, 285, 330, 365, 300, 310, 290, 295, 285, 360, 370, 275, 325, 335, 350, 280, 290)`.

-   **Instructions**:

    1.  Calculate the mean, median, and mode.

    2.  Calculate the variance and standard deviation.

    3.  Identify any outliers using the IQR method.

    4.  Write the R code to perform these calculations and interpret the results.

```{r}
# Sample data vector
reaction_times <- c(250, 340, 295, 305, 285, 330, 365, 300, 310, 290, 295, 285, 360, 370, 275, 325, 335, 350, 280, 290)

# Calculate mean


# Calculate median


# Calculate mode
get_mode <- function(x) {
  uniqv <- unique(x)
  uniqv[which.max(tabulate(match(x, uniqv)))]
}


# Calculate variance


# Calculate standard deviation


# Identify outliers using IQR


```

**Write your interpretation here**

### Exercise 2: Calculating Probabilities with the Normal Distribution

-   **Task**: Assume reaction times in a cognitive task follow a normal distribution with a mean of 300 ms and a standard deviation of 50 ms. Calculate the probability that a randomly selected individual has a reaction time:

    1.  Less than 250 ms.

    2.  Between 250 ms and 350 ms.

    3.  More than 400 ms.

-   **Instructions**:

    1.  Use the `pnorm` function in R to calculate these probabilities.

    2.  Write the R code for these calculations and explain what each probability means.

```{r}
# Parameters
mean <- 300
sd <- 50

# Probability of a reaction time less than 250 ms


# Probability of a reaction time between 250 ms and 350 ms


# Probability of a reaction time more than 400 ms


```

**Write your interpretation here**

### Exercise 3: Applying the T-Distribution

-   **Task**: You conducted a small-scale study with 8 participants measuring their anxiety levels on a scale of 1 to 10. Calculate the probability of a t-score being less than 2 and between -1 and 1 using the t-distribution.

-   **Instructions**:

    1.  Define the degrees of freedom for your study.

    2.  Use the `pt` function in R to calculate these probabilities.

    3.  Write the R code for these calculations and discuss how the results might differ if a normal distribution were assumed.

### Exercise 4: Defining and Simulating Sample Spaces

-   **Task**: Define a sample space for a study where participants choose their favorite relaxation activity (Reading, Music, Exercise, Meditation). Simulate responses from 40 participants.

-   **Instructions**:

    1.  Define the sample space.

    2.  Use the `sample` function in R to simulate the responses.

    3.  Analyze the frequency of each relaxation activity choice.

    4.  Write the R code to perform these tasks and summarize your findings.

```{r}
# Degrees of freedom
df <- 7  # for n = 8, df = n - 1

# Probability of a t-score less than 2


# Probability of a t-score between -1 and 1


```

**Write your interpretation here**


#### Submission Instructions:

Ensure to knit your document to PDF format, checking that all content is correctly displayed before submission. Submit this PDF to Canvas Assignments.
