---
title: "Weekly Lab Homework Assignment: Computation"
author: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

*Replace "Your Name" with your actual name.*

## Objective:

This lab assignment aims to reinforce your understanding of data cleaning and descriptive analysis using the `dplyr` and `psych` packages in R. You will apply these concepts through practical exercises, focusing on using and stacking `dplyr` functions with the `%>%` operator.

## Instructions:

1.  Complete each exercise by writing the necessary R code.

2.  Ensure you use the `%>%` operator to chain multiple `dplyr` functions together.

3.  Interpret the results for each exercise.

4.  Knit your R Markdown file to a PDF and submit it as per the submission instructions.

## Homework Exercises:

### Exercise 1: Cleaning Data with `dplyr`

Clean a dataset using various `dplyr` functions.

1.  Use the following dataset for the exercise:

```{r}
data <- data.frame(
  participant_id = 1:10,
  reaction_time = c(250, 340, 295, NA, 310, 275, 325, 290, 360, NA),
  gender = c("M", "F", "F", "M", "M", "F", "M", "F", "M", "F"),
  accuracy = c(95, 87, 92, 88, 94, 91, 85, 89, 93, NA)
)

print(data)
```

2.  Clean the dataset by performing the following steps:

    -   Remove rows with missing values.

    -   Rename the `reaction_time` column to `response_time`.

    -   Create a new column `performance_group` based on `accuracy` (High if `accuracy` \>= 90, otherwise Low).

    -   Remove outliers from the `response_time` column.

    -   Relevel the `performance_group` column to set "Low" as the reference level.

```{r, warning = F, message  F}
# Load necessary package
library(dplyr)

#create cleaned_data
cleaned_data <- data

#Now, with the cleaned_data object write code to:
# Remove rows with missing values
# Rename the reaction_time column
# Create a new column performance_group
# Remove outliers from the response_time column
# Relevel the performance_group column


# View the cleaned data
print(cleaned_data)

```

**Interpretation**: Describe the changes made to the dataset, such as the number of rows removed due to missing values, the new column created, and any outliers removed.

## Exercise 2: Generating Descriptive Statistics with `psych`

Generate descriptive statistics for a dataset.

1.  Use the following dataset for the exercise:

```{r}
study_hours <- data.frame(
  participant_id = 1:10,
  hours = c(5, 6, 4, 7, 5, 3, 8, 6, 5, 7)
)

```

2.  Generate descriptive statistics using the `describe()` function from the `psych` package.

```{r, warning = F, message = F}
#load the psych package
library(psych)

# Generate descriptive statistics


```

-   **Interpretation**: Explain the descriptive statistics obtained, such as the mean, standard deviation, skewness, and kurtosis of the `hours` variable.

## Exercise 3: Visualizing Data with `psych`

Create graphical summaries of a dataset using the `psych` package.

1.  Use the following dataset for the exercise:

```{r}
experiment_data <- data.frame(
  response_time = c(250, 340, 295, 310, 275, 325, 290, 360, 285, 310),
  accuracy = c(95, 87, 92, 88, 94, 91, 85, 89, 93, 90),
  age = c(23, 35, 29, 22, 30, 31, 27, 40, 24, 32)
)

```

-   Create a correlation plot using the `corPlot()` function.

```{r}
# Create the correlation plot


```

-   **Interpretation**: Describe the correlation coefficients displayed in the plot, indicating the strength and direction of relationships between variables.

-   Create pair panels using the `pairs.panels()` function.

```{r}
# Create the pair panels

```

**Interpretation**: Explain the scatterplots, histograms, and correlation coefficients in the pair panels, highlighting any notable patterns or relationships.

**Submission Instructions:**

Ensure to knit your document to PDF format, checking that all content is correctly displayed before submission. Submit this PDF to Canvas Assignments.
