---
title: "Weekly Lab Homework Assignment: Computation"
author: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

*Replace "Your Name" with your actual name.*

## Objective:

In this lab, you will apply data transformation techniques, including mean-centering, calculating Z-scores, and performing non-linear transformations on various datasets. Please complete the exercises by filling in the code chunks and answering the interpretation questions. Once completed, knit this document to PDF and submit it as instructed.

### Exercise 1: Mean-Centering

**Dataset**: - Simulated data on the number of hours spent studying per week: `r   study_hours <- c(15, 22, 18, 25, 20, 28, 24, 19, 23, 26)`

**Tasks**:\
1. Calculate the mean of the study hours.\
2. Mean-center the dataset by subtracting the mean from each value.\
3. Plot the original and mean-centered study hours on the same graph.\
4. **Interpretation**: Explain what the mean-centered values tell you about the amount of time each student spent studying compared to the average.

```{r}
study_hours <- c(15, 22, 18, 25, 20, 28, 24, 19, 23, 26)

# Calculate the mean of the study hours
mean_study_hours <- mean(study_hours)

# Mean-center the study hours
mean_centered_study_hours <- study_hours - mean_study_hours

# Plot the original and mean-centered study hours
y_limits <- range(c(study_hours, mean_centered_study_hours))
plot(study_hours, type = "b", col = "blue", ylab = "Study Hours", xlab = "Student", main = "Original vs Mean-Centered Study Hours", ylim = y_limits)
lines(mean_centered_study_hours, type = "b", col = "red")
legend("topright", legend = c("Original", "Mean-Centered"), col = c("blue", "red"), lty = 1)

```

**Interpretation:** The mean-centered values tell you how each student's study hours compare to the average study hours. A negative value indicates that the student studied fewer hours than average, while a positive value indicates that the student studied more hours than average. For example, a mean-centered value of -7 means that the student studied 7 hours less than the average of 22 hours.

### Exercise 2: Calculating Z-Scores

**Dataset**: - Simulated data on students' reaction times (in milliseconds): `r   reaction_times <- c(350, 420, 310, 390, 370, 450, 380, 340, 400, 360)`

**Tasks**:\
1. Calculate the mean and standard deviation of the reaction times.\
2. Compute the Z-scores for each reaction time.\
3. Plot the Z-scores on a line graph.\
4. **Interpretation**: Discuss what a Z-score greater than 0 or less than 0 indicates about a reaction time relative to the average.

```{r}
reaction_times <- c(350, 420, 310, 390, 370, 450, 380, 340, 400, 360)

# Calculate the mean and standard deviation of the reaction times
mean_reaction_times <- mean(reaction_times)
sd_reaction_times <- sd(reaction_times)

# Compute the Z-scores
z_scores_reaction_times <- (reaction_times - mean_reaction_times) / sd_reaction_times

# Plot the Z-scores
plot(z_scores_reaction_times, type = "b", col = "purple", ylab = "Z-Scores", xlab = "Student", main = "Z-Scores of Reaction Times")

```

**Interpretation:** A Z-score greater than 0 indicates that a reaction time is above average (slower), while a Z-score less than 0 indicates that a reaction time is below average (faster). Z-scores help standardize reaction times, making it easier to compare across different individuals. For example, a Z-score of 1.69 means that the reaction time was 1.69 standard deviations slower than the average, indicating a relatively slow response. Conversely, a Z-score of -1.55 indicates a faster-than-average response.

### Exercise 3: Non-Linear Transformations

**Dataset**: - Simulated data on annual sales figures (in thousands of dollars): `r   sales <- c(200, 450, 700, 1200, 300, 800, 1100, 900, 400, 1500)`

**Tasks**:\
1. Apply a logarithmic transformation to the sales data.\
2. Apply a square root transformation to the sales data.\
3. Plot histograms of the original and transformed sales data.\
4. **Interpretation**: Compare the distributions of the original and transformed data. Explain how each transformation affects the spread and shape of the data.

```{r}
sales <- c(200, 450, 700, 1200, 300, 800, 1100, 900, 400, 1500)

# Apply a logarithmic transformation
log_sales <- log(sales)

# Apply a square root transformation
sqrt_sales <- sqrt(sales)

# Plot histograms of the original and transformed sales data
par(mfrow = c(1, 3))
hist(sales, breaks = 10, col = "blue", xlab = "Sales", main = "Original Sales")
hist(log_sales, breaks = 10, col = "green", xlab = "Log(Sales)", main = "Log Transformed Sales")
hist(sqrt_sales, breaks = 10, col = "orange", xlab = "Sqrt(Sales)", main = "Square Root Transformed Sales")

```

**Interpretation:**

-   **Logarithmic Transformation:** The log transformation reduces the skewness of the sales data, making the distribution more balanced. It is particularly useful for right-skewed data with a few very large values.

-   **Square Root Transformation:** The square root transformation also reduces skewness and stabilizes variance, but to a lesser extent than the log transformation. This transformation is useful when there is less extreme skewness in the data.

-   Both transformations make the data more suitable for further analysis by reducing the influence of outliers and stabilizing variance across different ranges of data.

### Exercise 4: Combining Transformations

**Dataset**: - Simulated data on daily step counts: `r   step_counts <- c(8000, 10500, 9200, 11500, 10000, 12500, 11000, 9500, 10200, 12000)`

**Tasks**: 1. Mean-center the step counts.\
2. Calculate the Z-scores for the mean-centered step counts.\
3. Plot the original, mean-centered, and Z-scores on separate graphs.\
4. **Interpretation**: Explain how the combination of mean-centering and Z-scores helps in understanding the step count data compared to looking at the original data alone.

```{r}
step_counts <- c(8000, 10500, 9200, 11500, 10000, 12500, 11000, 9500, 10200, 12000)

# Mean-center the step counts
mean_step_counts <- mean(step_counts)
mean_centered_step_counts <- step_counts - mean_step_counts

# Calculate the Z-scores for the mean-centered step counts
sd_step_counts <- sd(step_counts)
z_scores_step_counts <- mean_centered_step_counts / sd_step_counts

# Plot the original, mean-centered, and Z-scores
par(mfrow = c(3, 1))
plot(step_counts, type = "b", col = "blue", ylab = "Step Counts", xlab = "Day", main = "Original Step Counts")
plot(mean_centered_step_counts, type = "b", col = "green", ylab = "Mean-Centered", xlab = "Day", main = "Mean-Centered Step Counts")
plot(z_scores_step_counts, type = "b", col = "red", ylab = "Z-Scores", xlab = "Day", main = "Z-Scores of Mean-Centered Step Counts")

# Interpretation: Provide your answer here
```

**Interpretation:** Mean-centering the step counts allows you to see how each day's step count compares to the average. Z-scores further standardize these mean-centered values, showing how many standard deviations each day's step count is from the mean. This combination of transformations makes it easier to identify patterns, such as which days had significantly higher or lower step counts compared to the overall average. For example, a Z-score of 1.55 indicates that the step count on that day was 1.55 standard deviations above the average, suggesting a significantly higher activity level.

**Submission Instructions:**

Ensure to knit your document to PDF format, checking that all content is correctly displayed before submission. Submit this PDF to Canvas Assignments.
