---
title: "Weekly Lab Homework Assignment: Hypothesis Testing for Samples from Two Populations"
author: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

*Replace "Your Name" with your actual name.*

### Instructions

In this lab assignment, you will apply your understanding of hypothesis testing by conducting various tests and interpreting their results. You'll work with different datasets to perform t-tests, calculate confidence intervals, and analyze the results. Make sure to follow the instructions carefully and complete all parts of each exercise. Remember to write your code in the provided code chunks and knit your R Markdown file to PDF before submitting it.

### Exercise 1: Conducting an Independent Samples t-Test

**Scenario**: You are studying the effect of two different teaching methods on student performance. You collect test scores from two groups of students who were taught using different methods. The data is as follows:

-   **Group A Scores**: `c(78, 82, 85, 88, 91, 77, 85, 89, 90, 92)`
-   **Group B Scores**: `c(70, 75, 80, 82, 84, 76, 78, 81, 83, 85)`

**Tasks**:

1\. Conduct an independent samples t-test to compare the means of the two groups using R.

2\. Interpret the t-value, degrees of freedom, and p-value from the output.

```{r}
# Sample data
group_A <- c(78, 82, 85, 88, 91, 77, 85, 89, 90, 92)
group_B <- c(70, 75, 80, 82, 84, 76, 78, 81, 83, 85)

# Conduct the independent samples t-test

```

-   **t-Value**:
-   **Degrees of Freedom**:
-   **p-Value**:
-   **Interpretation**:

### Exercise 2: Calculating Confidence Intervals

**Scenario**: You want to calculate a 95% confidence interval for the difference between the means of the two groups (Group A and Group B) from Exercise 1.

**Tasks**:

1\. Calculate the 95% confidence interval for the difference between the means of Group A and Group B using the t-test results.

2\. Interpret the confidence interval and discuss what it suggests about the difference between the two teaching methods.

```{r}
# Confidence interval from t-test result

```

-   **95% Confidence Interval**: [ ]
-   **Interpretation**:

### Exercise 3: Conducting a Paired Samples t-Test

**Scenario**: You collect data on the anxiety levels of participants before and after they complete a stress management program. The data is as follows:

-   **Anxiety Levels Before**: `c(60, 62, 65, 68, 70, 72, 74, 76, 78, 80)`
-   **Anxiety Levels After**: `c(55, 58, 60, 62, 63, 65, 66, 68, 70, 72)`

**Tasks**:

1\. Conduct a paired samples t-test to determine if there is a significant difference in anxiety levels before and after the program.

2\. Interpret the t-value, degrees of freedom, and p-value from the output.

```{r}
# Sample data
anxiety_before <- c(60, 62, 65, 68, 70, 72, 74, 76, 78, 80)
anxiety_after <- c(55, 58, 60, 62, 63, 65, 66, 68, 70, 72)

# Conduct the paired samples t-test

```

-   **t-Value**:
-   **Degrees of Freedom**:
-   **p-Value**:
-   **Interpretation**:

### Exercise 4: Understanding Significance and Effect Size

**Scenario**: You conduct a study comparing the effectiveness of two diets on weight loss. The independent samples t-test yields a p-value of 0.03 and a Cohen's d of 0.5.

**Tasks**:

1\. Discuss whether the result is statistically significant and what the p-value indicates.

2\. Interpret the effect size (Cohen's d = 0.5) and its practical significance in the context of the study's findings.

**Submission Instructions:**

Ensure to knit your document to PDF format, checking that all content is correctly displayed before submission. Submit this PDF to Canvas Assignments.
