---
title: "Weekly Lab Homework Assignment: Bivariate Models"
author: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

*Replace "Your Name" with your actual name.*

### Instructions

In this lab, you will practice creating and interpreting bivariate linear models using various datasets. For each exercise, follow the provided instructions to build a linear model, interpret the slope, and analyze the residuals. Remember to follow the format for interpreting the slope:

> ***"For every one unit increase in (X/IV), Y (Outcome/DV) increases/decreases by (SLOPE VALUE)."***

For the above format, it will be EITHER increases OR decreases depending on the sign of the slope (positive or negative).

### Exercise 1: Relationship Between Daily Exercise and Happiness

**Dataset**: You are provided with a dataset that includes the number of hours people spend exercising daily and their happiness scores on a scale from 0 to 100.

**Simulate the Data**:

```{r}
set.seed(101)
daily_exercise <- rnorm(100, mean = 1, sd = 0.5)  # Hours of daily exercise
happiness <- 50 + 5 * daily_exercise + rnorm(100, mean = 0, sd = 5)  # Happiness scores

# Combine into a data frame
exercise_data <- data.frame(daily_exercise, happiness)

# View the first few rows
head(exercise_data)
```

**Tasks**:

1\. **Create the Linear Model**: Build a linear model to predict happiness based on daily exercise.

2\. **Interpret the Slope**: Interpret the slope of the model using the provided format.

3\. **Plot the Residuals**: Analyze the residuals and discuss whether the linear model is appropriate.

```{r}
# Create the linear model
model_exercise <- lm(happiness ~ daily_exercise, data = exercise_data)

# View the summary of the model
summary(model_exercise)

# Interpret the slope
# The slope value is extracted from the summary output
slope_exercise <- coef(model_exercise)[2]

# Plot the residuals
residuals_exercise <- residuals(model_exercise)
plot(exercise_data$daily_exercise, residuals_exercise, 
     main = "Residuals Plot for Exercise vs. Happiness",
     xlab = "Daily Exercise (Hours)", ylab = "Residuals", pch = 19, col = "blue")
abline(h = 0, col = "red", lwd = 2)
```

**Interpretation**: - Interpret the slope using the format: "For every one unit increase in study time, academic performance increases/decreases by [SLOPE VALUE]."

### Exercise 2: Relationship Between Screen Time and Sleep Quality

**Dataset**: You have data on the number of hours of screen time before bed and sleep quality scores (0 to 100).

**Simulate the Data**:

```{r}
set.seed(102)
screen_time <- rnorm(100, mean = 3, sd = 1)  # Hours of screen time before bed
sleep_quality <- 80 - 4 * screen_time + rnorm(100, mean = 0, sd = 8)  # Sleep quality scores

# Combine into a data frame
sleep_data <- data.frame(screen_time, sleep_quality)

# View the first few rows
head(sleep_data)
```

**Tasks**:

1\. **Create the Linear Model**: Build a linear model to predict sleep quality based on screen time.

2\. **Interpret the Slope**: Interpret the slope of the model using the provided format.

3\. **Plot the Residuals**: Analyze the residuals and discuss whether the linear model is appropriate.

```{r}
# Create the linear model
model_sleep <- lm(sleep_quality ~ screen_time, data = sleep_data)

# View the summary of the model
summary(model_sleep)

# Interpret the slope
# The slope value is extracted from the summary output
slope_sleep <- coef(model_sleep)[2]

# Plot the residuals
residuals_sleep <- residuals(model_sleep)
plot(sleep_data$screen_time, residuals_sleep, 
     main = "Residuals Plot for Screen Time vs. Sleep Quality",
     xlab = "Screen Time (Hours)", ylab = "Residuals", pch = 19, col = "purple")
abline(h = 0, col = "red", lwd = 2)
```

**Interpretation**: - Interpret the slope using the format: "For every one unit increase in study time, academic performance increases/decreases by [SLOPE VALUE]."


### Exercise 3: Relationship Between Coffee Consumption and Productivity

**Dataset**: You are given data on the number of cups of coffee consumed daily and productivity scores at work (0 to 100).

**Simulate the Data**:

```{r}
set.seed(103)
coffee_consumption <- rpois(100, lambda = 3)  # Number of cups of coffee
productivity <- 60 + 2.5 * coffee_consumption + rnorm(100, mean = 0, sd = 7)  # Productivity scores

# Combine into a data frame
coffee_data <- data.frame(coffee_consumption, productivity)

# View the first few rows
head(coffee_data)
```

**Tasks**:

1\. **Create the Linear Model**: Build a linear model to predict productivity based on coffee consumption.

2\. **Interpret the Slope**: Interpret the slope of the model using the provided format.

3\. **Plot the Residuals**: Analyze the residuals and discuss whether the linear model is appropriate.

```{r}
# Create the linear model
model_coffee <- lm(productivity ~ coffee_consumption, data = coffee_data)

# View the summary of the model
summary(model_coffee)

# Interpret the slope
# The slope value is extracted from the summary output
slope_coffee <- coef(model_coffee)[2]

# Plot the residuals
residuals_coffee <- residuals(model_coffee)
plot(coffee_data$coffee_consumption, residuals_coffee, 
     main = "Residuals Plot for Coffee Consumption vs. Productivity",
     xlab = "Cups of Coffee", ylab = "Residuals", pch = 19, col = "green")
abline(h = 0, col = "red", lwd = 2)
```

**Interpretation**: - Interpret the slope using the format: "For every one unit increase in study time, academic performance increases/decreases by [SLOPE VALUE]."


### Exercise 4: Relationship Between Social Media Usage and Loneliness

**Dataset**: This dataset includes information on the number of hours spent on social media daily and participants' loneliness scores (0 to 100).

**Simulate the Data**:

```{r}
set.seed(104)
social_media <- rnorm(100, mean = 2, sd = 1)  # Hours of social media use
loneliness <- 40 + 7 * social_media + rnorm(100, mean = 0, sd = 6)  # Loneliness scores

# Combine into a data frame
social_media_data <- data.frame(social_media, loneliness)

# View the first few rows
head(social_media_data)
```

**Tasks**:

1\. **Create the Linear Model**: Build a linear model to predict loneliness based on social media usage.

2\. **Interpret the Slope**: Interpret the slope of the model using the provided format.

3\. **Plot the Residuals**: Analyze the residuals and discuss whether the linear model is appropriate.

```{r}
# Create the linear model
model_social_media <- lm(loneliness ~ social_media, data = social_media_data)

# View the summary of the model
summary(model_social_media)

# Interpret the slope
# The slope value is extracted from the summary output
slope_social_media <- coef(model_social_media)[2]

# Plot the residuals
residuals_social_media <- residuals(model_social_media)
plot(social_media_data$social_media, residuals_social_media, 
     main = "Residuals Plot for Social Media Usage vs. Loneliness",
     xlab = "Social Media Usage (Hours)", ylab = "Residuals", pch = 19, col = "orange")
abline(h = 0, col = "red", lwd = 2)
```

**Interpretation**: - Interpret the slope using the format: "For every one unit increase in study time, academic performance increases/decreases by [SLOPE VALUE]."


### Exercise 5: Relationship Between Study Time and Academic Performance

**Dataset**: This dataset includes data on the number of hours students study per week and their academic performance scores (0 to 100).

**Simulate the Data**:

```{r}
set.seed(105)
study_time <- rnorm(100, mean = 5, sd = 2)  # Hours of study per week
academic_performance <- 70 + 3 * study_time + rnorm(100, mean = 0, sd = 10)  # Academic performance scores

# Combine into a data frame
study_data <- data.frame(study_time, academic_performance)

# View the first few rows
head(study_data)
```

**Tasks**:\
1. **Create the Linear Model**: Build a linear model to predict academic performance based on study time.

2\. **Interpret the Slope**: Interpret the slope of the model using the provided format.

3\. **Plot the Residuals**: Analyze the residuals and discuss whether the linear model is appropriate.

```{r}
# Create the linear model
model_study <- lm(academic_performance ~ study_time, data = study_data)

# View the summary of the model
summary(model_study)

# Interpret the slope
# The slope value is extracted from the summary output
slope_study <- coef(model_study)[2]

# Plot the residuals
residuals_study <- residuals(model_study)
plot(study_data$study_time, residuals_study, 
     main = "Residuals Plot for Study Time vs. Academic Performance",
     xlab = "Study Time (Hours per Week)", ylab = "Residuals", pch = 19, col = "red")
abline(h = 0, col = "blue", lwd = 2)
```

**Interpretation**: - Interpret the slope using the format: "For every one unit increase in study time, academic performance increases/decreases by [SLOPE VALUE]."

**Submission Instructions:**

Ensure to knit your document to PDF format, checking that all content is correctly displayed before submission. Submit this PDF to Canvas Assignments.
