---
title: "Weekly Lab Homework Assignment: Correlations"
author: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

*Replace "Your Name" with your actual name.*

### Instructions

In this lab assignment, you will apply what you've learned about correlation by calculating, visualizing, and interpreting correlations using different datasets. Make sure to follow the instructions carefully and complete all parts of each exercise. Remember to write your code in the provided code chunks and knit your R Markdown file to PDF before submitting it.

### Exercise 1: Calculating Pearson Correlation Coefficient

**Scenario**: You are examining the relationship between the number of hours spent exercising per week and self-reported levels of happiness. The data is as follows:

-   **Exercise Hours**: `c(1, 3, 4, 6, 8, 10)`
-   **Happiness Scores**: `c(50, 55, 60, 65, 70, 75)`

**Tasks**:

1\. Calculate the Pearson correlation coefficient using R.

2\. Interpret the correlation coefficient in the context of the relationship between exercise hours and happiness scores.

```{r}
# Sample data
exercise_hours <- c(1, 3, 4, 6, 8, 10)
happiness_scores <- c(50, 55, 60, 65, 70, 75)

# Calculate Pearson's correlation coefficient

```

-   **Calculated Pearson Correlation Coefficient**: 0.996
-   **Interpretation**: The correlation coefficient of 0.996 indicates a very strong positive correlation between exercise hours and happiness scores. This suggests that as exercise hours increase, happiness scores increase in a closely related manner.

### Exercise 2: Visualizing Correlation with ggplot2

**Scenario**: You are investigating the relationship between daily water intake (in liters) and energy levels throughout the day. The data is as follows:

-   **Water Intake (liters)**: `c(0.5, 1, 1.5, 2, 2.5, 3)`
-   **Energy Levels**: `c(40, 50, 60, 65, 70, 80)`

**Tasks**:

1\. Create a scatter plot using ggplot2 in R to visualize the relationship between water intake and energy levels.

2\. Add a trend line to the scatter plot to show the direction and strength of the correlation.

**R Code Chunk**:

```{r, warning = F}
library(ggplot2)

# Sample data
water_intake <- c(0.5, 1, 1.5, 2, 2.5, 3)
energy_levels <- c(40, 50, 60, 65, 70, 80)

# Create scatter plot with trend line

```

-   **Interpretation**: The scatter plot with a trend line shows a clear positive relationship between water intake and energy levels. The upward-sloping trend line indicates that higher water intake is associated with higher energy levels, consistent with the likely positive correlation.

### Exercise 3: Analyzing the Size of Correlation

**Scenario**: You have calculated the Pearson correlation coefficient for two variables: screen time per day and sleep duration per night. The correlation coefficient is $r = -0.4$.

**Tasks**:

1\. Interpret the size of the correlation.

2\. Discuss the practical significance of this correlation in the context of the relationship between screen time and sleep duration.

### Exercise 4: Impact of a Third Variable (Confounder)

**Scenario**: You observe a correlation between outdoor time and academic performance in a group of students. However, you suspect that physical health might be influencing both.

**Tasks**:

1\. Discuss how physical health could act as a confounding variable affecting both outdoor time and academic performance.

2\. Suggest methods to control for this third variable in future research.

### Exercise 5: Evaluating Correlation and Causality

**Scenario**: A study finds a strong positive correlation between eating breakfast and overall cognitive performance in children. However, the study does not examine causality.

**Tasks**:

1\. Discuss why the correlation between eating breakfast and cognitive performance does not necessarily imply that eating breakfast causes better cognitive performance.

2\. Provide specific examples of how you could design a study to explore causality between these variables.

**Submission Instructions:**

Ensure to knit your document to PDF format, checking that all content is correctly displayed before submission. Submit this PDF to Canvas Assignments.
